Computing Direct Sum Decompositions

### 2025-04-28 paper/main.tex
devlin's final edits

The problems of finding isomorphism classes of indecomposable modules with a given property, or determining the indecomposable summands of a module, are ubiquitous in commutative algebra and representation theory and even in areas such as topological data analysis. Within commutative algebra, for instance, the classification of rings $R$ for which there are only finitely many isomorphism classes of indecomposable maximal Cohen--Macaulay $R$-modules (the \emph{finite CM-type} property), or determining whether iterated Frobenius pushforwards of a Noetherian ring in positive characteristic have finitely many isomorphism classes of indecomposable summands (the \emph{finite F-representation type} property) are two well-established research problems (see, for example, \cite{Yoshino90,LW12} for the former and \cite{SVdB97,Hara15,TT08}, among many others, for the latter). For both these problems, and many others, making and testing conjectures depends on computing summands of modules and verifying their indecomposability.

Currently there are no algorithms available for performing these tasks on an arbitrary module over a commutative ring. The need for and absence of such algorithms is noted, for example, in \cite[\S15.10.9]{Eisenbud95} and \cite[\S8.5]{BL23}. In contrast, variants of the ``Meat-Axe'' algorithm for determining irreducibility of finite-dimensional modules over a group algebra have wide ranging applications in computational group theory \cite{Parker84,HR94,Holt98,IL00} and are available through symbolic algebra software such as Magma and GAP \cite{MAGMA,GAP}.

@@

The purpose of this paper is to describe and prove correctness of two practical algorithms for computing indecomposable summands of finitely generated modules over a $k$-algebra $R$: \Cref{alg:graded} if $M$ is a homogeneous module over a (multi)graded ring $R$ with $k$ a field of arbitrary characteristic, and \Cref{alg:local} if $R$ is local and $k$ a field of positive characteristic. Over a large enough field, our algorithms are often capable of completely decomposing a module in a single non-recursive iteration.

@@

Let $R$ be a $\N$-graded ring with $R_0 = k$ a field. Suppose $M = \bigoplus [M]_d$ is a finitely generated, graded $R$-module and consider a general degree-zero endomorphism $\phi \in [\End_R(M)]_0$. Fix a set of minimal homogeneous generators $m_1,\dots,m_{\mu(M)}$ of $M$ so that $\phi$ may be presented as a $\mu(M)\times\mu(M)$ matrix with entries in $R$. We will refer to $\chi_\phi(\lambda) = \det(\phi-\lambda\id_M)$ as the ``characteristic polynomial of $\phi$'', which is a univariate polynomial with coefficients in $R$ and thus a priori has solutions only in the algebraic closure of the total ring of fractions of $R$.

\begin{prop}\label{prop:eigenvalues}
  Let $A\colon M/\m M\to M/\m M$ be the $k$-linear map induced by $\phi\in[\End_R(M)]_0$. The solutions of the characteristic polynomial $\chi_\phi(\lambda)$ are the same as the eigenvalues of $A$, thus they lie in the algebraic closure $\bar k$.
\end{prop}
\begin{proof}
  Since $M$ is $\N$-graded, assume its generators are ordered with $\deg m_i$ nonincreasing, and let $d_1,\dots,d_n$ be the decreasing list of unique degrees of generators of $M$. The matrix for $\phi$ is then ``block upper triangular'' with $n$ diagonal blocks with entries in $k$, i.e., of the form
  \begin{equation}\label{eq:blocks}
    \phi = \begin{pmatrix}
    B_1    & *      & \dots  & *      \\
    0      & B_2    & \dots  & *      \\
    \vdots & \vdots & \ddots & \vdots \\
    0      & 0      & \dots  & B_n
    \end{pmatrix},
  \end{equation}
  where the blocks $B_i$ are square matrices \emph{with entries in $k$}, and the remaining entries are in $\m$. To see this, note that since $\phi$ has degree zero, $\phi(m_i)$ is a $k$-linear combination of the $m_j$ with $\deg m_j = \deg m_i$ plus an $R$-linear combination of the $m_j$ with $\deg m_j < \deg m_i$. %In particular, $B_i$ is an $k$-endomorphism on the $k$-vector space spanned by all generators of degree $d_i$.

  Then the characteristic polynomial $\det(\phi - \lambda \id_M) = \prod \det(B_i - \lambda) = \det(A - \lambda \id_{M/\m M})$ is a univariate polynomial in $\lambda$ with coefficients in $k$, thus has solutions in $\bar k$.
\end{proof}

\begin{dfn}\label{def:eigenvalues}
  The collection of eigenvalues of $B_i$ in the splitting field of $\chi_\phi(\lambda)$ are the \emph{eigenvalues of $\phi$}. We will use $\mu_j \coloneqq \mu(\lambda_j)$ for the sum of the (geometric) multiplicities of $\lambda_j$ in each $B_i$. Note that the (algebraic) multiplicity of $\lambda_j$ in $\chi_\phi(\lambda)$ is the sum of the multiplicities of $\lambda_j$ in each $\chi_{B_i}(\lambda)$ and may be larger than $\mu_j$.
\end{dfn}

\begin{rem}\label{rem:grading}
  The hypotheses that $R$ is $\N$-graded may be weakened. In the proof of \Cref{prop:eigenvalues}, we only need a partial order on the degrees of the generators of $M$ in order to present $\phi$ in a block upper triangular form \eqref{eq:blocks}. Suppose $R$ is $\ZZ^r$-graded, then the effective cone $\Eff(R)\subset\Q^n$ generated by the elements $\deg m$ for each monomial $m \in R$ induces a partial order on the degrees of generators of $M$ if it is strongly convex (contains no positive dimensional subspace of $\Q^n$) and $R_0 = k$. We will call such rings \emph{positively graded}. % TODO: can R_0 be finite dimensional?
  This is satisfied, for instance, in the setting of multiparameter persistence modules or when $R$ is the Cox ring of a toric variety or Mori dream space graded by the class group of the variety.
\end{rem}

@@

Note that as long as each $\psi_i$ in \Cref{lem:decomposition} is nontrivial (not zero or identity endomorphisms), the kernel summands will be nontrivial, however, they may not necessarily be indecomposable. Also note that $\im\psi$ may be zero.

\begin{prop}\label{prop:split-surj}
  Let $\lambda_1,\dots,\lambda_r$ be the eigenvalues of $\phi$ over $k$ and set $\psi_j = (\phi - \lambda_j\id_M)^{\mu_j}$ and $\psi = \psi_1\circ\cdots\circ\psi_r$. Then $M$ has a direct sum decomposition
  \( M = \ker\psi_1 \oplus \cdots \oplus \ker\psi_r \oplus \im\psi. \)
\end{prop}
\begin{proof}
  If $\psi_i$ is injective on $\im\psi_i$, then $\im\psi_i$ is a summand: in this case, the composition $\im\psi_i\hookrightarrow M\to \im\psi_i$ is injective, but a degree-zero endomorphism of a finitely generated $R$-module is an isomorphism (it must be injective, hence surjective, on each finite-dimensional $[M]_d$), and thus the inclusion $\im \psi \to M$ is split by the surjection $M\to \im \psi$. Therefore the short exact sequence $0\to\ker\psi_i\to M\to\im\psi_i\to 0$ is split and $M = \ker\psi_i\oplus\im\psi_i$.

  To see that $\psi_i$ is injective on $\im\psi_i$, first for simplicity replace $\phi$ by $\phi-\lambda_i\id_M$ so $\psi_i = \phi^{\mu_i}$. By the Cayley--Hamilton theorem $\psi_i$ divides $\chi_\phi(\phi)$, hence
  \begin{align*}
    \chi_\phi(\phi) = \left(\phi^{\mu(M) - \mu_i} - \dots + c_1\phi - c_0\right) \circ \psi_i
    % &= \left(\phi^{\mu(M) - \mu_i} - \dots + c_1\phi\right) \circ \psi_i - c_0 \psi_i \\
    &= \left(\phi^{\mu(M) - \mu_i - 1} - \dots + c_1\right) \circ \phi \circ \psi_i - c_0 \psi_i = 0,
  \end{align*}
  for some $c_0,c_1$ in $k$, and therefore
  \( \psi_i = c_0^{-1} \left(\phi^{\mu(M) - \mu_i - 1} - \dots + c_1\right) \circ \phi \circ \psi_i. \)
  This means that $\phi$ is injective on $\im\psi_i$, therefore $\psi_i=\phi^{\mu_i}$ is injective on $\im\psi_i$.

  Moreover, using the standard proof for vector spaces, we know $\ker\psi_i\cap\ker\psi_j = 0$ if $i\neq j$: suppose $(\phi - \lambda_i)^m(v) = 0$ for some $v\in M$ and $m$ is minimal so that $w = (\phi-\lambda_i)^{m-1}(v) \neq 0$ satisfies $(\phi-\lambda_i)(w) = 0$, but then since $(\phi - \lambda_j)(\phi - \lambda_i) = (\phi - \lambda_i)(\phi - \lambda_j)$ we get
  \[
  (\phi - \lambda_j)(w)
  = (\phi - \lambda_j)(\phi - \lambda_i)^{m-1}(v)
  = (\phi - \lambda_i)^{m-1}(\phi - \lambda_j)(v) = 0,
  \]
  which implies
  \(
  0 = (\phi - \lambda_j)(w) = (\phi - \lambda_i + \lambda_i - \lambda_j)(w) = (\lambda_i - \lambda_j)(w),
  \)
  which is a contradiction.

@@

\begin{algorithm}[H]
  \caption{(Indecomposable summands of a graded module over a commutative ring)}\label{alg:graded}
  \begin{algorithmic}[1]
    \smallskip
    \Require graded module $M$ over a positively graded commutative $k$-algebra $R$.
    \Ensure  list $\textproc{Summands}(M)$ of indecomposable summands of $M$.
    \State Initialize a list $L$.
    \State Take a general element $\phi$ in $[\End_R(M)]_0$,
           the degree zero part of $\End_R(M)$. \label{item:End0}
    \ForAll { eigenvalues $\lambda_i$ of $\phi$ }
      \State Let $\mu_i$ be the (geometric) multiplicity of $\lambda_i$.
      \State Let $\psi_i = (\phi - \lambda_i \id_M)^{\mu_i}$ (or, for simplicity, raised to $\mu(M)$).
      \State By \Cref{prop:split-surj}, $\psi_i$ is a split surjection, hence producing a splitting of $M$.
      \State Append \Call{Summands}{$\ker\psi_i$} to $L$.
    \EndFor
    \State Let $\psi$ be the composition of all $\psi_i$ above.
    \If {$\psi$ is nonzero}
      \State Append \Call{Summands}{$\im\psi$} to $L$.
    \EndIf
    \State \Return $L$.
  \end{algorithmic}
\end{algorithm}

@@

\begin{rem}\label{rem:graded}
  Note that \Cref{alg:graded} does not depend on the characteristic of the field. Further, since over a large enough field (i.e., when $|k|$ is larger than the number of summands) the eigenvalues of a general endomorphism $\phi$ are distinct (e.g., by a variant of \Cref{lem:distinct}), the summands $\ker\psi_i$ and $\im\psi$ will be indecomposable; thus \Cref{alg:graded} will produce an indecomposable decomposition of $M$ in one non-recursive iteration. When $|k|$ is less than or approximately the same as the number of summands, the algorithm will be recursive.
\end{rem}

@@

\begin{lem}\label{lem:idemp}
  Let $A\colon M/\m M\to M/\m M$ be the induced action of $\psi\in\End_R(M)$ on $M/\m M$. If $A$ is an idempotent, then $M$ admits a direct sum decomposition $\ker\psi \oplus \im\psi$.
\end{lem}
\begin{proof}
  Let $N \coloneqq \im\psi$. We want to show that $0 \to N \inc M$ splits. Consider the composition
  \begin{equation}\label{eq:N-M-N}
    N \inc M \xra{\psi} N.
  \end{equation}
  We claim that this composition is surjective. Since a surjective endomorphism of finitely generated modules is invertible, we conclude that this composition is an isomorphism $\alpha$ on $N$.
  Therefore the inclusion $ 0 \to N \inc M $ is split by the composition
  \[ M \xra{\psi} N \xra{\alpha\inv} N, \]
  and thus $M$ decomposes as claimed.

@@

  By the Artin--Rees lemma \cite[Lemma~5.1]{Eisenbud95}, there is a number $\ell>0$ such that for any $i\gg0$ we get
  \[ \m^i M\cap N = \m^{i-\ell} ( \m^i M\cap N)\subset \m^{i-\ell} N. \]
  That is, the terms of $n_0-n_1+n_2-\cdots$ are going to 0 in the $\m$-adic topology on $N$. Hence we can write
  \[ n_0 = \psi(n_0-n_1+n_2-\cdots), \]
  with $n_0-n_1+n_2-\cdots\in N$. We conclude that $\psi$ is surjective as a map $N\to N$.
\end{proof}

Thus, if we produce an element $\psi \in \End_R(M)$ that is an idempotent modulo $\m$, we obtain a splitting of $M$.
The following lemma allows us to produce idempotents modulo $\m$.

\begin{lem}\label{lem:jordan}
  Let $k$ be a finite field of characteristic $p$, and let $A$ be an endomorphism of a $k$-vector space such that all eigenvalues of $A$ are contained in $k$. If $\lambda$ is an eigenvalue of $A$, then some power of $A-\lambda$ is idempotent.
  Furthermore, if $\lambda$ is not the only eigenvalue of $A$, then a  nonzero power of $A-\lambda$ is idempotent.
\end{lem}
\begin{proof}
  Since all eigenvalues of $A$ are contained in $k$, we can without loss of generality put $A$ in Jordan canonical form, with each Jordan block being an $r_i\times r_i$ matrix of the form
  \[ \begin{pmatrix}
        \lambda_i & 1 & 0 & \dots & 0 \\
        0 & \lambda_i & 1  & \dots & 0 \\
        %0 & 0 & \lambda_i & \dots & 0 \\
        \vdots & \vdots & \vdots & \ddots & \vdots \\
        0 & 0 & 0 & \dots & \lambda_i
    \end{pmatrix}, \]
  where each $\lambda_i$ is an eigenvalue of $A$.
  In this basis, $A-\lambda$ will be block-diagonal with blocks
  \[ \begin{pmatrix}
    \lambda_i-\lambda & 1 & 0 & \dots & 0 \\
    0 & \lambda_i-\lambda & 1 & \dots & 0 \\
    \vdots & \vdots & \vdots & \ddots & \vdots \\
    0 & 0 & 0 & \dots & \lambda_i-\lambda
  \end{pmatrix}. \]
  Set $\nu_i=\lambda_i-\lambda$. Then for any $n\geq 1$,
  the $n$-th power $(A-\lambda)^n$ is block-diagonal with blocks
  \[ \begin{pmatrix}
    \nu_i^n & \binom{n}1\nu_i^{n-1} & \binom{n}2\nu_i^{n-2} & \dots & \binom{n}{r_i}\nu_i^{n-r_i} \\
    0 & \nu_i^n & \binom{n}1\nu_i^{n-1} & \dots & \binom{n}{r_i-1}\nu_i^{n-r_i+1} \\
    \vdots & \vdots & \vdots & \ddots & \vdots \\
    0 & 0 & 0 & \dots & \nu_i^n
  \end{pmatrix}. \]
  If we choose $n > r_i$ and $n$ a power of $p$, say $p^{e_0}$, then all non-diagonal terms will vanish, so all blocks will have the form
  \[ \begin{pmatrix}
    \nu_i^n & 0 & 0 & \dots & 0 \\
    0 & \nu_i^n & 0 & \dots & 0 \\
    \vdots & \vdots & \vdots & \ddots & \vdots \\
    0 & 0 & 0 & \dots & \nu_i^n
  \end{pmatrix}. \]
  Finally, since $k\subset \FF_{p^e}$ for some $e$, if we choose $n$ to be divisible also by $p^e-1$, we have that $$\nu_i^{n-1} = (\nu_i^{p^{e}-1})^{n/(p^{e}-1)} =1.$$
  In particular, taking $n=p^{e_0}(p^e-1)$, the matrix $(A-\lambda)^{p^{e_0}(p^e-1)}$ is a diagonal matrix with diagonal entries 1 or 0, hence idempotent. Note moreover that if some $\lambda_i\neq \lambda$, then $(A-\lambda)^n$ is not the zero matrix.
\end{proof}

\begin{prop}\label{prop:split-idems}
  Let $\lambda_1,\dots,\lambda_r$ be the eigenvalues of $\phi\otimes R/\m$ defined over $\FF_{p^e}$ and set $\psi_j = (\phi - \lambda_j \id_M)^{p^{e_i} (p^{e} - 1)}$ with $e_i = \lceil \log_p(\mu_i) \rceil$ and $\psi = \psi_1\circ\cdots\circ\psi_r$. Then $M$ has a direct sum decomposition
  \( M = \ker\psi_1 \oplus \cdots \oplus \ker\psi_r \oplus \im\psi. \)
\end{prop}
\begin{proof}
  The proof is analogous to \Cref{prop:split-surj}. Specifically, using \Cref{lem:jordan} and \Cref{lem:idemp}, each $\psi_i$ is an idempotent, and in particular, a split surjection. Further, $\ker\psi_i\cap\ker\psi_j = 0$ by the same argument as in the proof of \Cref{prop:split-surj}. Therefore, using \Cref{lem:decomposition}, we have the desired direct sum decomposition.
\end{proof}

>>>>

The problems of finding isomorphism classes of indecomposable modules with a given property, or determining the indecomposable summands of a module, are ubiquitous in commutative algebra and representation theory and even in areas such as topological data analysis. Within commutative algebra, for instance, the classification of rings $R$ for which there are only finitely many isomorphism classes of indecomposable maximal Cohen--Macaulay $R$-modules (the \emph{finite CM-type} property), or for which iterated Frobenius pushforwards in positive characteristic have finitely many isomorphism classes of indecomposable summands (the \emph{finite F-representation type} property) are two well-established research problems (see, for example, \cite{Yoshino90,LW12} for the former and \cite{SVdB97,Hara15,TT08}, among many others, for the latter). For both these problems, and many others, making and testing conjectures depends on computing summands of modules and verifying their indecomposability.

Currently there are no algorithms available for finding direct sum decompositions of an arbitrary module over a commutative ring. The need for and absence of such algorithms is noted, for example, in \cite[\S15.10.9]{Eisenbud95} and \cite[\S8.5]{BL23}. In contrast, variants of the ``Meat-Axe'' algorithm for determining irreducibility of finite-dimensional modules over a group algebra have wide-ranging applications in computational group theory \cite{Parker84,HR94,Holt98,IL00} and are available through symbolic algebra software such as Magma and GAP \cite{MAGMA,GAP}.
\looseness-1

@@

The purpose of this paper is to describe and prove correctness of two practical algorithms for computing indecomposable summands of finitely generated modules over a $k$-algebra $R$: \Cref{alg:graded} in the case that $M$ is a homogeneous module over a (multi)graded ring $R$ with $k$ a field of arbitrary characteristic, and \Cref{alg:local} when $R$ is local and $k$ a field of positive characteristic contained in $\bar \F_p$. Over a large enough field, our algorithms are often capable of completely decomposing a module in a single non-recursive iteration.

@@

Let $R$ be a $\N$-graded ring with $R_0 = k$ a field. Suppose $M = \bigoplus [M]_d$ is a finitely generated, graded $R$-module and consider a  degree-zero endomorphism $\phi \in [\End_R(M)]_0$. Fix a set of minimal homogeneous generators $m_1,\dots,m_{\mu(M)}$ of $M$ so that $\phi$ may be presented as a $\mu(M)\times\mu(M)$ matrix with entries in $R$. We will refer to $\chi_\phi(\lambda) = \det(\phi-\lambda\id_M)$ as the ``characteristic polynomial of $\phi$'', which is a univariate polynomial with coefficients in $R$ and thus a priori has solutions only in the algebraic closure of the total ring of fractions of $R$.

\begin{prop}\label{prop:eigenvalues}
  Let $A\colon M/\m M\to M/\m M$ be the $k$-linear map induced by $\phi\in[\End_R(M)]_0$. The solutions of the characteristic polynomial $\chi_\phi(\lambda)$ are the same as the eigenvalues of $A$, thus they lie in the algebraic closure $\bar k$.
\end{prop}
\begin{proof}
  Since $M$ is $\N$-graded, we may order its generators such that $\deg m_i$ is nonincreasing; let $d_1,\dots,d_n$ be the decreasing list of unique degrees of generators of $M$. The matrix for $\phi$ is then ``block upper triangular'' with $n$ diagonal blocks with entries in $k$, i.e., of the form
  \begin{equation}\label{eq:blocks}
    \phi = \begin{pmatrix}
    B_1    & *      & \dots  & *      \\
    0      & B_2    & \dots  & *      \\
    \vdots & \vdots & \ddots & \vdots \\
    0      & 0      & \dots  & B_n
    \end{pmatrix},
  \end{equation}
  where the blocks $B_i$ are square matrices \emph{with entries in $k$}, and the remaining entries are in $\m$. To see this, note that since $\phi$ has degree zero, $\phi(m_i)$ is a $k$-linear combination of the $m_j$ with $\deg m_j = \deg m_i$ plus an $\m$-linear combination of the $m_j$ with $\deg m_j < \deg m_i$. %In particular, $B_i$ is an $k$-endomorphism on the $k$-vector space spanned by all generators of degree $d_i$.

  Thus the characteristic polynomial $\det(\phi - \lambda \id_M) = \prod \det(B_i - \lambda) = \det(A - \lambda \id_{M/\m M})$ is a univariate polynomial in $\lambda$ with coefficients in $k$, thus has solutions in $\bar k$.
\end{proof}

\begin{dfn}\label{def:eigenvalues}
  The collection of eigenvalues of $B_i$ in the splitting field of $\chi_\phi(\lambda)$ are the \emph{eigenvalues of $\phi$}. We will write $\mu_j \coloneqq \mu(\lambda_j)$ for the sum of the (geometric) multiplicities of $\lambda_j$ in each $B_i$. Note that the (algebraic) multiplicity of $\lambda_j$ in $\chi_\phi(\lambda)$ is the sum of the multiplicities of $\lambda_j$ in each $\chi_{B_i}(\lambda)$ and may be larger than $\mu_j$.
\end{dfn}

\begin{rem}\label{rem:grading}
  The hypotheses that $R$ is $\N$-graded may be weakened. In the proof of \Cref{prop:eigenvalues}, we only need a partial order on the degrees of the generators of $M$ in order to present $\phi$ in a block upper triangular form \eqref{eq:blocks}. If $R$ is $\ZZ^r$-graded, then the \emph{effective cone} $\Eff(R)\subset\Q^n$ generated by the elements $\deg m$ for each monomial $m \in R$ induces a partial order on the degrees of generators of $M$ if it is strongly convex (contains no positive dimensional subspace of $\Q^n$) and $R_0 = k$. We will call such rings \emph{positively graded}. % TODO: can R_0 be finite dimensional?
  This is satisfied, for instance, in the setting of multiparameter persistence modules or when $R$ is the Cox ring of a toric variety or Mori dream space graded by the class group of the variety.
\end{rem}

@@

Note that as long as each $\psi_i$ in \Cref{lem:decomposition} is nontrivial (not the zero or identity endomorphism), the kernel summands will be nontrivial (although they may not necessarily be indecomposable). Also note that $\im\psi$ may be zero.

\begin{prop}\label{prop:split-surj}
  Let $\lambda_1,\dots,\lambda_r$ be the eigenvalues of $\phi$ over $k$ and set $\psi_j = (\phi - \lambda_j\id_M)^{\mu_j}$ and $\psi = \psi_1\circ\cdots\circ\psi_r$. Then $M$ has a direct sum decomposition
  \( M = \ker\psi_1 \oplus \cdots \oplus \ker\psi_r \oplus \im\psi. \)
\end{prop}
\begin{proof}
  If $\psi_i$ is injective on $\im\psi_i$, then $\im\psi_i$ is a summand: in this case, the composition $\im\psi_i\hookrightarrow M\to \im\psi_i$ is injective, but a degree-zero endomorphism of a finitely generated $R$-module is an isomorphism (it must be injective, hence surjective, on each finite-dimensional $[M]_d$). Thus, the inclusion $\im \psi \to M$ is split by the surjection $M\to \im \psi$. Therefore the short exact sequence $0\to\ker\psi_i\to M\to\im\psi_i\to 0$ is split and $M = \ker\psi_i\oplus\im\psi_i$.

  To see that $\psi_i$ is injective on $\im\psi_i$, first for simplicity replace $\phi$ by $\phi-\lambda_i\id_M$ so $\psi_i = \phi^{\mu_i}$. By the Cayley--Hamilton theorem (see, e.g., \cite[Theorem~4.3]{Eisenbud95}), $\psi_i$ divides $\chi_\phi(\phi)$, hence
  \begin{align*}
    \chi_\phi(\phi) = \left(\phi^{\mu(M) - \mu_i} - \dots + c_1\phi - c_0\right) \circ \psi_i
    % &= \left(\phi^{\mu(M) - \mu_i} - \dots + c_1\phi\right) \circ \psi_i - c_0 \psi_i \\
    &= \left(\phi^{\mu(M) - \mu_i - 1} - \dots + c_1\right) \circ \phi \circ \psi_i - c_0 \psi_i = 0,
  \end{align*}
  for some $c_0,c_1$ in $k$, and therefore
  \( \psi_i = c_0^{-1} \left(\phi^{\mu(M) - \mu_i - 1} - \dots + c_1\right) \circ \phi \circ \psi_i. \)
  This means that $\phi$ is injective on $\im\psi_i$, and therefore the composition $\psi_i=\phi^{\mu_i}$ is still injective on $\im\psi_i$.
We thus have that $\psi_i$ is a split surjection, as desired.

  To see that $\ker\psi_i\cap\ker\psi_j = 0$ if $i\neq j$, we mimic the standard proof for vector spaces: suppose $(\phi - \lambda_i)^m(v) = 0$ for some $v\in M$ and $m$ is minimal, so that $w = (\phi-\lambda_i)^{m-1}(v) \neq 0$ satisfies $(\phi-\lambda_i)(w) = 0$. 
If $(\phi - \lambda_j)^{\mu_j}(v) = 0$ as well, then since $(\phi - \lambda_j)(\phi - \lambda_i) = (\phi - \lambda_i)(\phi - \lambda_j)$ (as $\phi$ commutes with itself and the identity) we get
  \[
  (\phi - \lambda_j)^{\mu_j}(w)
  = (\phi - \lambda_j)^{\mu_j}(\phi - \lambda_i)^{m-1}(v)
  = (\phi - \lambda_i)^{m-1}(\phi - \lambda_j)^{\mu_j}(v) = 0.
  \]
Thus,
  \(
  0 = (\phi - \lambda_j)^{\mu_j}(w) = (\phi - \lambda_i + \lambda_i - \lambda_j)^{\mu_j}(w) = (\lambda_i - \lambda_j)^{\mu_j}(w)
  \)
(using that $(\phi - \lambda_i)(w)=0$).
Since $\lambda_i-\lambda_j \in k$ is nonzero, 
$(\lambda_i - \lambda_j)^{\mu_j}(w)= 0 $ forces $w=0$, a contradiction.

@@

\begin{algorithm}[H]
  \caption{(Indecomposable summands of a graded module over a commutative ring)}\label{alg:graded}
  \begin{algorithmic}[1]
    \smallskip
    \Require graded module $M$ over a positively graded commutative $k$-algebra $R$.
    \Ensure  list $\textproc{Summands}(M)$ of indecomposable summands of $M$.
    \State Initialize a list $L$.
    \State Take a general element $\phi$ in $[\End_R(M)]_0$,
           the degree-zero part of $\End_R(M)$. \label{item:End0}
    \ForAll { eigenvalues $\lambda_i$ of $\phi$ }
      \State Let $\mu_i$ be the (geometric) multiplicity of $\lambda_i$.
      \State Let $\psi_i = (\phi - \lambda_i \id_M)^{\mu_i}$ (or, for simplicity, raised to $\mu(M)$).
      \State By \Cref{prop:split-surj}, $\psi_i$ is a split surjection, hence producing a splitting of $M$.
      \State Append \Call{Summands}{$\ker\psi_i$} to $L$.
    \EndFor
    \State Let $\psi$ be the composition of all $\psi_i$ above.
    \If {$\psi$ is nonzero}
      \State Append \Call{Summands}{$\im\psi$} to $L$.
    \EndIf
    \State \Return $L$.
  \end{algorithmic}
\end{algorithm}

@@

\begin{rem}\label{rem:graded}
  Note that \Cref{alg:graded} does not depend on the characteristic of the field. Further, since over a large enough field (i.e., when $|k|$ is larger than the number of summands) the eigenvalues of a general endomorphism $\phi$ are are as distinct as  possible (by a variant of \Cref{lem:distinct}), the summands $\ker\psi_i$ and $\im\psi$ will be indecomposable; thus \Cref{alg:graded} will produce an indecomposable decomposition of $M$ in one non-recursive iteration. When $|k|$ is less than or approximately the same as the number of summands, the algorithm will be recursive.
\looseness-1
\end{rem}

@@

\begin{lem}\label{lem:idemp}
  Let $A\colon M/\m M\to M/\m M$ be the induced action of $\psi\in\End_R(M)$ on $M/\m M$. If $A$ is an idempotent, then $M$ admits a direct sum decomposition $\ker\psi \oplus \im\psi$.
\end{lem}
\begin{proof}
  Let $N \coloneqq \im\psi$. We want to show that $0 \to N \inc M$ splits. Consider the composition
  \begin{equation}\label{eq:N-M-N}
    N \inc M \xra{\psi} N.
  \end{equation}
  We claim that this composition is surjective. Since a surjective endomorphism of finitely generated modules is invertible, we conclude that this composition is an isomorphism $\alpha$ on $N$.
  Therefore the inclusion $ 0 \to N \inc M $ is split by the composition
  \( M \xra{\psi} N \xra{\alpha\inv} N, \)
  and thus $M$ decomposes as claimed.

@@

  By the Artin--Rees lemma \cite[Lemma~5.1]{Eisenbud95}, there is $\ell>0$ such that for $i\gg0$ we get
  \[ \m^i M\cap N = \m^{i-\ell} ( \m^i M\cap N)\subset \m^{i-\ell} N. \]
  That is, the terms of $n_0-n_1+\cdots$ go to 0 in the $\m$-adic topology on $N$. Thus we can write
  \[ n_0 = \psi(n_0-n_1+n_2-\cdots), \]
  with $n_0-n_1+n_2-\cdots\in N$. We conclude that $\psi$ is surjective as a map $N\to N$.
\end{proof}

Thus, if we produce an element $\psi \in \End_R(M)$ that is an idempotent modulo $\m$, we obtain a splitting of $M$.
The following lemma allows us to produce idempotents modulo $\m$.
\vadjust{\goodbreak}

\begin{lem}\label{lem:jordan}
  Let $k\subset \bar \F_{p}$, and let $A$ be an endomorphism of a $k$-vector space. 
Choose:
\begin{itemize}
\item $e_0$ such that $p^{e_0}$ is larger than the geometric multiplicity of any eigenvalue of $A$ in $\bar k$.
\item $e$ is such that all eigenvalues of $A$ in $\bar k$ lie in $\F_{p^e}$.
\end{itemize}
If $\lambda$ is an eigenvalue of $A$ contained in $k$, then $(A-\lambda)^{p^{e_0} (p^{e}-1)}$ is idempotent.
  If $\lambda$ is not the only eigenvalue of $A$ over $\bar k$, this is a nontrivial idempotent.
\end{lem}
\begin{proof}
To check that this power is idempotent, we may extend to the algebraic closure $\bar k$, because if $A^2=A$ as an endomorphism of $\bar k$-vector spaces the same is true over $k$.
  Since all eigenvalues of $A$ are contained in $\bar k$, we can without loss of generality put $A$ in Jordan canonical form, with each Jordan block being an $r_i\times r_i$ matrix of the form
  \[ \begin{pmatrix}
        \lambda_i & 1 & 0 & \dots & 0 \\
        0 & \lambda_i & 1  & \dots & 0 \\
        %0 & 0 & \lambda_i & \dots & 0 \\
        \vdots & \vdots & \vdots & \ddots & \vdots \\
        0 & 0 & 0 & \dots & \lambda_i
    \end{pmatrix}, \]
  where each $\lambda_i$ is an eigenvalue of $A$ and $r_i$ is bounded above by the geometric multiplicity of $\lambda_i$.
  In this basis, $A-\lambda$ will be block-diagonal with blocks
  \[ \begin{pmatrix}
    \lambda_i-\lambda & 1 & 0 & \dots & 0 \\
    0 & \lambda_i-\lambda & 1 & \dots & 0 \\
    \vdots & \vdots & \vdots & \ddots & \vdots \\
    0 & 0 & 0 & \dots & \lambda_i-\lambda
  \end{pmatrix}. \]
  Set $\nu_i=\lambda_i-\lambda$. Then for any $n\geq 1$,
  the $n$-th power $(A-\lambda)^n$ is block-diagonal with blocks
  \[ \begin{pmatrix}
    \nu_i^n & \binom{n}1\nu_i^{n-1} & \binom{n}2\nu_i^{n-2} & \dots & \binom{n}{r_i}\nu_i^{n-r_i} \\
    0 & \nu_i^n & \binom{n}1\nu_i^{n-1} & \dots & \binom{n}{r_i-1}\nu_i^{n-r_i+1} \\
    \vdots & \vdots & \vdots & \ddots & \vdots \\
    0 & 0 & 0 & \dots & \nu_i^n
  \end{pmatrix}. \]
  If we choose $n > r_i$ and a power of $p$, for example, $p^{e_0}$, then all non-diagonal terms will vanish, so all blocks will have the form
  \[ \begin{pmatrix}
    \nu_i^n & 0 & 0 & \dots & 0 \\
    0 & \nu_i^n & 0 & \dots & 0 \\
    \vdots & \vdots & \vdots & \ddots & \vdots \\
    0 & 0 & 0 & \dots & \nu_i^n
  \end{pmatrix}. \]
  Finally, since $\nu_i =\lambda-\lambda_i \in \FF_{p^e}$ for each $i$, if we choose $n$ to be divisible also by $p^e-1$, we have that $$\nu_i^{n} = (\nu_i^{p^{e}-1})^{n/(p^{e}-1)} =1.$$
  In particular, taking $n=p^{e_0}(p^e-1)$, the matrix $(A-\lambda)^{p^{e_0}(p^e-1)}$ is a diagonal matrix with diagonal entries 1 or 0, hence idempotent. Note moreover that if some $\lambda_i\neq \lambda$, then $(A-\lambda)^n$ is not the zero matrix.
\end{proof}

\begin{prop}\label{prop:split-idems}
  Let $\lambda_1,\dots,\lambda_r$ be the eigenvalues of $\phi\otimes R/\m$ defined over $k$; pick $e$ such that all eigenvalues of $\phi \otimes R/\m$ in $\bar k$ are contained in $\F_{p^e}$
and $e_0 =  \lceil \log_p(\mu(M))\rceil+1$.
Set $\psi_j = (\phi - \lambda_j \id_M)^{p^{e_0} (p^{e} - 1)}$ 
and $\psi = \psi_1\circ\cdots\circ\psi_r$. Then $M$ has a direct sum decomposition
  \( M = \ker\psi_1 \oplus \cdots \oplus \ker\psi_r \oplus \im\psi. \)
\end{prop}

We note that $e$ can be calculated effectively by computing the splitting field of $\chi(\phi\otimes R/\m)$.

\begin{proof}
  The proof is analogous to the proof of \Cref{prop:split-surj}. By \Cref{lem:jordan} and \Cref{lem:idemp},
since $p^{e_0}> \mu(M)$ is an upper bound on the geometric multiplicity of any eigenvalue in $\bar k$,
 each $\psi_i$ is idempotent and in particular a split surjection. Further, $\ker\psi_i\cap\ker\psi_j = 0$ by the same argument as in the proof of \Cref{prop:split-surj}. Therefore, using \Cref{lem:decomposition}, we have the desired direct sum decomposition.
\end{proof}
